Synthesis of Linear Functional Programs

### 2021-07-18 article/article.tex
WIP: Introduction to focusing, grammars...

\usepackage{cmll}
\usepackage{mathpartir}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{hyperref}

\usepackage{minted}

\usepackage{todonotes}

\usepackage{multirow}

@@

\newcommand{\llet}[2]{\mathsf{let}\,#1\,\mathsf{in}\,#2}

@@

\paragraph{Outline.}
First, we discuss linear types as specifications, the outline of the
synthesis process, and what that entails for the synthetiser -- ilustrated with
concrete results (Section~\ref{sec:overview}). Then, we present the
algorithm and formal system driving the synthesis
(Section~\ref{sec:formal_system}). Aftwerwards, we describe the architecture 
of our synthetiser named \synname, and examine technical details and challenges of the
implementation (Section~\ref{sec:architecture}). Finally, we evaluate our work
through expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks
(i.e.~how fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
%TODO: And appendice with...

@@

\todo[inline]{Aqui parece-me que deves comecar por explicar o que vais
  fazer nesta seccao. Sinto falta da sintaxe da linguagem,
  possivelmente incrementalmente. Comecava por listar os tipos/termos
  do lambda linear ``simples'', depois em cada subseccao acrescentava
  as construcoes novas e como a sintese funciona para esse
  contexto. Enfatizar que a novelty é a combinacao destas features todas.}

The development of \synname\ was based on an almost direct reading of
concrete inference rules and theoretical principles (~\cite{frank
  pfenning notes});
\todo{aqui queres dizer que começaste por te basear
  num sistema de focusing para a logica linear (era sequer linear a
  formulacao que comecaste a seguir?) O que são theoretical
  principles?}
and work on its environment was also guided by formal
systems and methods (e.g.~Hindley-Milner type
inference~\cite{}).\todo{esta frase parece-me desnecessaria. O que
  queres dizer neste paragrafo, parece-me, é que partes de um sistema
  focused da literatura que depois vais extender, inspirando te em
  tecnicas standard de proof search / inferencia.}

However, with the addition of recursion and
richer types to the synthesis and language, formal guidelines
\todo{nao sao formal guidelines, nao existem trabalhos a combinar
  estas features todas, essa parte é importante. Além de que 
ao incluir estas coisas deixamos de estar completamente em proof
search, pois muitas destas coisas (e.g. recursao) nao sao
necessariamente sound logicamente.
  Tenta usar termos 
  especificos. Richer types? Richer como?} to support
the implementation become scarcer, or more complex \todo{nao existem
  ou sao complicados? ambos? citacao?}. Accordingly,
rather than derived from explicit rules, solutions appear first to
tame the extensions' inherent complex problems (such as infinite
recursion), and implicit rules come only as a byproduct. \todo{ Nao
  percebo o que queres dizer aqui. O que sao explicit rules e implicit
  rules? Ou há regras ou nao há. O que estas a dizer é que inventaste
  tecnicas de sintese para estes casos, que correspondem as regras que
  vais apresentar. E tao explicito ou implicito como o que esta na literatura.}

With this section we aim to formalize the rules that guide
the synthesis process for more expressive specifications, alongside the already
well defined rules that model the core of the synthetiser, resulting in
a \emph{sound} set of rules that characterizes our framework for synthesis of
recursive functions from polymorphic and refinement types in a linear context
% TODO: Better?....
, and describes the system in enough detail for
the synthetiser to be reproducible by a theory-driven implementation.

@@

\mypara{Core Rules} The core system comprises of the sequent calculus rules for
linear logic connectives with resource-management and focusing.
\todo{Introdução para focusing...}
Focusing separates the proof search into two
phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later try to
invert, moving them to the linear context ($\Delta$) when we fail to.

\todo{Primeiro explica os varios judgments (as ``formas'' das
  conclusoes das regras) e o que significam as varias componentes,
  nomeadamente a parte da gestao de recursos.  Depois podes explicar
  regras concretas como fazes a seguir. Tambem seria util dizer quais
  os tipos que sao invertiveis a esquerda/direita, pois o leitor nao
  faz ideia do que estas a dizer.}

Putting together linear logic, the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synthetise a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

The decision rules are followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synthetise recursive
programs from more expressive (richer) types.

\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
is a named sum type of any type, i.e. a named type that can be instantiated by one of many
tags (also known as constructors) that take exactly one fixed type, which might
be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
for constructors with an arbitrary number of parameters. In the \synname\
language, the programmer can define custom ADTs; as an example, we show the
definition of an ADT which might be constructed with zero, one, or two
propositions of type $A$, using its syntax: \mintinline{haskell}{data Container
= None 1 | One A | Two (A * A)}

The semantics of the ADTs relate to those of the plus ($\oplus$) type --
both are additive disjunctions.  A proof of an ADT requires only a proof of one
of the types that a constructor takes, similar to the way $\oplus$ requires only
proof of either the left or right type it consists of. Analogously, a proof of
$C$ by deconstructing an ADT stipulates that all types that can construct the
ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
the inference rules for a simple ADT mimic the $\oplus$ rules, for an
arbitrary ($> 1$) number of types, rather than exactly two. Therefore, there's one
left rule for ADTs, and an arbitrary number of right rules, one for each
constructor:

@@

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning constructors can take as arguments
values of the type they are defining. Seemingly small, this change has a big
impact in the synthesis process. Take, for instance, the ADT defined as
\mintinline{haskell}{data T = C1 T}, the synth goal $T \lolli C$, and its
derivation:
%
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {
                    \infer*[right=(adtL)]
                    {\dots}
                    {\Gamma; \Delta/\Delta'; \Omega, z{:}T \Uparrow\ \vdash \textrm{case}\ z\ 
                    \textrm{of}\ C_1\ k \rightarrow \dots : C}
                }
                {\Gamma; \Delta/\Delta'; \Omega, y{:}T \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of}\ C_1\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of}\ C_1\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}T \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x . \dots :
    T \lolli C \Uparrow}
\end{mathpar}
%
Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Symmetrically, the
derivation for the synth goal $T$ is also infinite.
%
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {
                \infer*[left=(adtR)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
            }
            {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
\end{mathpar}
%
To account for this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and, symetrically, don't allow construction of an
ADT when trying to synthetise an argument for its constructor.
%
%
For this, we need two more contexts, $\Rho_C$
for constraints on construction and $\Rho_D$ for constraints on
deconstruction. Together, they hold a list of ADTs that can not be constructed or
deconstructed at some point in the proof. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:
%
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    T \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        T \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    T\ \textrm{\textbf{else}}\ \Rho_C

@@

\mypara{Memoization} Some proofs attempt to synthetise the same sub-goal with
the same premises more than once. Since these proof attempts might be expensive
performance-wise, and we can rely on the program's determinism (equal calls to
synthetise result in equal outcomes), we add to our synthetiser memoization, an
optimization technique that stores function calls and returns cached values for
equal inputs. For our synthetiser, two calls are equivalent when contexts
$\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and
the goal, are equivalent. Because comparing all of these is slow, our
implementation hashes the relevant values and uses the result as the key.
Furthermore, a branch we want to memoize might fail, not providing any result --
we also want to record synthesis failure, as it happens more often than
successful results. We capture failure of a logict computation using the $ifte$
function and represent failure or success information with the Maybe type. We
ran benchmarks to measure the impact of memoization on the performance. It's
especially noticeable in synthesis with multiple functions in the unrestricted
context, or more complex specifications:
%
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Goal & Time + $\sigma$ & TimeM + $\sigma$ \\
        \hline
        insertWithKey & 13.31 s $\pm$ 593.4 ms & 2.667 s $\pm$ 284.3 ms \\
        \hline
    \end{tabular}
\end{center}
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality for polymorphic types.

@@

\mypara{Program Synthesis from Polymorphic Refinement Types} The
work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis
recursive functional programs but in a more ``advanced'' setting. The
specification is a combination of two rich forms of types: polymorphic and
refinement types (which correspond to a first-order logic through the
Curry-Howard isomorphism) -- this is an interesting combination that offers
``expressive power and decidability, which enables automatic
verification--and hence synthesis--of nontrivial programs''.  Their
approach to refinement types consists a new algorithm that supports
decomposition of the refinement specification, allowing for separation
between the language of specification and programs and making the
approach amenable to compositional synthesis techniques.

>>>>

\usepackage{cmll}
\usepackage{mathpartir}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{minted}
\usepackage{todonotes}
\usepackage{multirow}

@@

\newcommand{\llet}[2]{\mathsf{let}\ #1\ \mathsf{in}\ #2}

@@

\paragraph{Outline.} First, we discuss linear types as specifications, the
outline of the synthesis process, and what that entails for the synthetiser --
ilustrated with concrete results (Section~\ref{sec:overview}). Then, we present
the formal system driving the synthesis (Section~\ref{sec:formal_system}).
Aftwerwards, we describe the architecture of our synthetiser named \synname, and
examine technical details and challenges of the implementation
(Section~\ref{sec:architecture}). Finally, we evaluate our work through
expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks (i.e.~how
fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
%TODO: And appendice with...

@@

%\todo[inline]{Aqui parece-me que deves comecar por explicar o que vais fazer
%nesta seccao. Sinto falta da sintaxe da linguagem, possivelmente
%incrementalmente. Comecava por listar os tipos/termos do lambda linear
%``simples'', depois em cada subseccao acrescentava as construcoes novas e como a
%sintese funciona para esse contexto. Enfatizar que a novelty é a combinacao
%destas features todas.}

%The development of \synname\ was based on an almost direct reading of concrete
%inference rules and theoretical principles (~\cite{frank pfenning notes});
%%
%\todo{aqui queres dizer que começaste por te basear num sistema de focusing para
%a logica linear (era sequer linear a formulacao que comecaste a seguir?) O que
%são theoretical principles?}
%%
%and work on its environment was also guided by formal systems and methods
%(e.g.~Hindley-Milner type inference~\cite{}).
%%
%\todo{esta frase parece-me desnecessaria. O que queres dizer neste paragrafo,
%parece-me, é que partes de um sistema focused da literatura que depois vais
%extender, inspirando te em tecnicas standard de proof search / inferencia.}

%However, with the addition of recursion and richer types to the synthesis and
%language, formal guidelines
%%
%\todo{nao sao formal guidelines, nao existem trabalhos a combinar estas features
%todas, essa parte é importante. Além de que ao incluir estas coisas deixamos de
%estar completamente em proof search, pois muitas destas coisas (e.g. recursao)
%nao sao necessariamente sound logicamente.  Tenta usar termos especificos.
%Richer types? Richer como?}
%%
%to support the implementation become scarcer, or more complex.
%%
%\todo{nao existem ou sao complicados? ambos? citacao?}
%%
%Accordingly, rather than derived
%from rules, solutions appeared first as techniques to tame the extensions'
%inherent complex problems (such as infinite recursion), and we now present
%those techniques as rules.
%%
% \todo{ Nao percebo o que queres dizer aqui. O que sao explicit rules e implicit rules?
% Ou há regras ou nao há. O que estas a
% dizer é que inventaste tecnicas de sintese para estes casos, que correspondem
% as regras que vais apresentar. E tao explicito ou implicito como o que esta na
% literatura.}

Program synthesis from linear types in combination with polymorphism, recursive
algebraic data types, and refinements, is new in the synthesis 
literature. Despite the substantial amount of research on linear logic and
proof-search upon which we found our core synthetiser, formal guidelines for the
richer types and their inherent complex problems (such as infinite recursion)
don't exist. % TODO: Developed an algorithm...?

With this section we aim to formalize as rules the novel techniques that guide
synthesis from our more expressive specifications, alongside the already well
defined rules that model the core of the synthetiser, putting together a
\emph{sound} set of inference rules that characterizes our framework for synthesis of
recursive programs from specifications with the select richer types in a linear context,
% TODO: Better?....
and describes the system in enough detail for the synthetiser to be reproducible
by a theory-driven implementation.

@@

\mypara{Core Rules} The core system comprises of proof-search in sequent
calculus rules for linear logic with resource-management and focusing.
%
Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove \emph{inessencial} non-determinism from
proof-search by structuring the application of invertible and non-invertible
inference rules. Rules are classified as \emph{left} or \emph{right},
\emph{invertible} or \emph{non-invertible} as follows:...\todo{não estou a
conseguir formular bem esta introdução}
The proof search is separated into two phases:
%
inversion ($\Uparrow$), in which we apply all invertible rules eagerly, and
focusing ($\Downarrow$), in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later
try to invert, moving them to the linear context ($\Delta$) when we fail to.

% TODO: Explicar vários judgments?
% \todo{Primeiro explica os varios judgments (as ``formas'' das
%   conclusoes das regras) e o que significam as varias componentes,
%   nomeadamente a parte da gestao de recursos.  Depois podes explicar
%   regras concretas como fazes a seguir. Tambem seria util dizer quais
%   os tipos que sao invertiveis a esquerda/direita, pois o leitor nao
%   faz ideia do que estas a dizer.}
%
% Esta parte em geral está a ficar pouco coerente, com a introdução do focusing,
  % a suposta introdução dos judgments, e a introdução da gramática... deixa

The core language is a simply typed linear lambda calculus for which the terms
($M,N$)
and types' ($\tau$) syntax is constructed with the following grammar:
\[
\begin{tabular}{lclc}
$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
    N_2)$ & $\oplus$ \\
    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
    \\
$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
\end{tabular}
\]
%
Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synthetise a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

The decision rules are followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    { }
    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synthetise recursive
programs from more expressive (richer) types.

\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
is a named sum type of any type, i.e. a named type that can be instantiated by one of many
tags (also known as constructors) that take exactly one fixed type, which might
be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
for constructors with an arbitrary number of parameters. In the \synname\
language, the programmer can define custom ADTs; as an example, we show the
definition of an ADT which might be constructed with zero, one, or two
propositions of type $A$, using its syntax: \mintinline{haskell}{data Container
= None 1 | One A | Two (A * A)}. The language grammar is extended as:
\[
\begin{tabular}{lclc}
    $M,N$ & $\ ::=\ $ & $\dots\ \vert\ \emph{Cons}_n\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of}\ \dots\ \vert\ \emph{Cons}_n\ u \Rightarrow N)$ \\
    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ \langle\emph{name}\rangle$ \\
    $ADT$ & $\ ::=\ $ & (\textsf{data} $\langle\emph{name}\rangle$\ $=$\ $\emph{Cons}_1\ \tau$\
$\vert$\ $\dots$\ $\vert$\ $\emph{Cons}_n\ \tau$) \\ \end{tabular}
\]
%TODO: Como definir gramática dos ADTs?

The semantics of the ADTs relate to those of the plus ($\oplus$) type --
both are additive disjunctions.  A proof of an ADT requires only a proof of one
of the types that a constructor takes, similar to the way $\oplus$ requires only
proof of either the left or right type it consists of. Analogously, a proof of
$C$ by deconstructing an ADT stipulates that all types that can construct the
ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
the inference rules for a simple ADT mimic the $\oplus$ rules, for an
arbitrary ($>1$) number of types, rather than exactly two. Therefore, there's one
left rule for ADTs, and an arbitrary number of right rules, one for each
constructor:

@@

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning constructors can take as arguments
values of the type they are defining. Seemingly small, this change has a big
impact in the synthesis process. Take, for instance, the ADT defined as
\mintinline{haskell}{data T = C1 T}, the synth goal $T \lolli C$, and its
derivation:
%
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega, y{:}T \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of}\ C_1\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of}\ C_1\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}T \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x . \dots :
    T \lolli C \Uparrow}
\end{mathpar}
%
Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Symmetrically, the
derivation for the synth goal $T$ is also infinite.
%
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {\dots}
            {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
\end{mathpar}
%
To account for this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and, symetrically, don't allow construction of an
ADT when trying to synthetise an argument for its constructor.
%
%
For this, we need two more contexts, $\Rho_C$
for constraints on construction and $\Rho_D$ for constraints on
deconstruction. Together, they hold a list of ADTs that can not be constructed or
deconstructed at some point in the proof. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:
%
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    T \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        T \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    T\ \textrm{\textbf{else}}\ \Rho_C

@@

\mypara{Memoization} Some proofs attempt to synthetise the same sub-goal with
the same premises more than once. Since these proof attempts might be expensive
performance-wise, and we can rely on the program's determinism (equal calls to
synthetise result in equal outcomes), we add to our synthetiser memoization, an
optimization technique that stores function calls and returns cached values for
equal inputs. For our synthetiser, two calls are equivalent when contexts
$\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and
the goal, are equivalent. Because comparing all of these is slow, our
implementation hashes the relevant values and uses the result as the key.
Furthermore, a branch we want to memoize might fail, not providing any result --
we also want to record synthesis failure, as it happens more often than
successful results. We capture failure of a logict computation using the $ifte$
function and represent failure or success information with the Maybe type. We
ran benchmarks to measure the impact of memoization on the performance. It's
especially noticeable in synthesis with multiple functions in the unrestricted
context, or more complex specifications:
%TODO: ignore constraints if type isnnt existential
%
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Goal & Time + $\sigma$ & TimeM + $\sigma$ \\
        \hline
        insertWithKey & 13.31 s $\pm$ 593.4 ms & 2.667 s $\pm$ 284.3 ms \\
        \hline
    \end{tabular}
\end{center}
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality for polymorphic types.

@@

\mypara{Program Synthesis from Polymorphic Refinement Types} The
work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis
recursive functional programs but in a more ``advanced'' setting. The
specification is a combination of two rich forms of types: polymorphic and
refinement types (which correspond to a first-order logic through the
Curry-Howard isomorphism) -- this is an interesting combination that offers
``expressive power and decidability, which enables automatic
verification--and hence synthesis--of nontrivial programs''.  Their
approach to refinement types consists of a new algorithm that supports
decomposition of the refinement specification, allowing for separation
between the language of specification and programs and making the
approach amenable to compositional synthesis techniques.
